\documentclass[../../proposal.tex]{subfiles}

\begin{document}

The goal of this research is to establish the role of lightweight formal methods as an invaluable tool in bridging the gap between modern software engineering practices and the unique challenges presented by scientific and engineering software.  This involves the following specific objectives:

\begin{enumerate}

    \item To clearly define common paradigms of lightweight formal methods within the context of scientific software development.  These include but are not limited to: specification, verification and validation, correctness, refinement, and predicate abstraction.
    
    \item To demonstrate the utility of lightweight formal methods through the development of actual models and their respective software.  These examples will reflect concepts commonly found in scientific and engineering software, and will span a broad range of abstractions in order to demonstrate the utility and applicability of the approach.  Examples are to include the moment distribution method, sparse matrix multiplication, the finite element method, web-based analysis tools, and user interface design.

    \item To develop tools that aid scientists and engineers in applying these methods in practice.  These tools include the a web-based domain specific visualization tool for use in the modeling of finite element software, the development of a model sharing utility that leverages version control and the visualization tool, and the embedding of the visualization tool and the Alloy analyzer into an existing IDE (integrated development environment).

\end{enumerate}

\end{document}